The Hyperedge Event Model

### 2018-05-26 paper/ba-HEM/ba-sample.tex
finished revising BA-HEM draft

	We evaluate the model's ability to predict edges and timestamps from Montgomery County email data, conditioned on their ``training" part of the data. To perform the experiment, we separately formed a test split of each three model components---sender, receivers, and timestamps---by randomly selecting ``test" data with probability $p=0.1$. Any missing variables were imputed by drawing samples from their conditional posterior distributions, given the observed data, model estimates, and current values of test data. We then run inference to update the latent variables given the imputed and observed data. We iterate the two steps---imputation and inference---multiple times to obtain enough number of estimates for test data. Algorithm \ref{alg:PPE} outlines this procedure in detail. We run the experiment and measure the predictive performance of two separate time distributions using $N=500$ predicted samples, by comparing the predictions in terms of classification accuracy in predicting the senders and receivers, as well as prediction error in the timestamps. First, we compare the posterior probability of correct senders for each of missing emails $\{d:a_d=\mbox{NA}\}$, which corresponds to $\pi_{a_{d}}=P(a_{d} = a^{obs}_{d}|\cdot)$ in Algorithm \ref{alg:PPE}. We call this measure as ``correct sender posterior probability." On the left of Figure \ref{figure:PPEresults}, we draw boxplots for the distribution of mean correct sender posterior probability---i.e., $\hat{\pi}_{a_{d}} = \frac{1}{N}\sum_{n=1}^N \pi^n_{a_{d}}$---across the missing emails. The results show that both log-normal and exponential distributions acheive better predictive performance for missing senders compared to what is expected under random guess (i.e., choose one out of $A$ possible senders $=1/18$), with log-normal distribution showing slightly higher performance than exponential distribution on average. Secondly, since the receiver vector is binary, we compute $F_1$ scores for missing receiver indicators (i.e., all $d$ and $r$ with $r_{dr}$=NA) by taking the harmonic mean of precision and recall:

@@

	with TP denoting true positive (i.e., $\boldsymbol{r}^{obs}_{dr}=\boldsymbol{r}^{pred}_{dr}=1$), FN denoting false negative (i.e., $\boldsymbol{r}^{obs}_{dr}=1$ but $\boldsymbol{r}^{pred}_{dr}=0$), and FP denoting false positive (i.e., $\boldsymbol{r}^{obs}_{dr}=0$ but $\boldsymbol{r}^{pred}_{dr}=1$). Although the generative process for edges (Section \ref{subsec: Tie}) is not directly affected by the choice of timestamp distribution, the middle plot in Figure \ref{figure:PPEresults} reveals noticeable difference between log-normal and exponential in their performance in predicting missing receiver indicators, where log-normal outperforms exponential. Finally, the prediction error for $d^{th}$ missing timestamp is estimated using the median of absolute relative errors\footnote{\cite{hyndman2006another} refer to this as median absolute percentage error (MdAPE).} across $N=500$ predictions:

@@

			\caption {Comparison of predictive performance between log-normal and exponential distributions: boxplots for correct sender posterior probability (\textit{left}), $F_1$ scores from receiver predictions (\textit{middle}), and boxplots for median absolute percentage error (\textit{right}). The dotted blue lines in the left plot represents the correct sender probability expected by random guess---i.e., $1/A=1/18\approx0.056$.}

@@

@@

	The right plot of Figure \ref{figure:PPEresults} presents boxplots for the median absolute relative errors. where plot the estimates in a log-scale. Surprisingly, we have a huge benefit in the performance of timestamp prediction when we assume log-normal distribution for time-increments compared to exponential distribution. This difference can be explained by overdispersion in exponential distribution, because there exists greater variability in the time increments of emails than would be expected under exponential distribution. As illustrated above, we can use this out-of-sample prediction task for two usage---1) provide an effective answer to the question ``how are we filling in the missing components of time-stamped network data?" and 2) offer one standard way to determine the distribution of time increments in Section \ref{subsec:Time}.

@@

	\subsubsection{Coefficients for edge covariates}
	Figure \ref{figure:betaresults} shows the boxplots summarizing posterior samples of $\boldsymbol{b}$, where (a) displays the coefficients for nodal covariates and (b) displays the coefficients for dyadic and triadic covariates. Since we use the logit probability of $\lambda_{adr}$ when generating hypothetical edges in Section \ref{subsec: Tie}, we assume that 
	\begin{equation*}
	\mbox{logit}(\lambda_{adr})=\log\Big(\frac{\lambda_{adr}}{1-\lambda_{adr}}\Big) =b_{1}+b_{2} x_{adr2}\ldots+b_{14}x_{adr14},
	\end{equation*}
	and can interpret the $\boldsymbol{b}$ estimates in terms of odds ratio $\frac{\lambda_{adr}}{1-\lambda_{adr}}=\exp(b_{1}+b_{2} x_{adr2}\ldots+b_{14}x_{adr14})$.
	To begin with, we find out that the effect of nodal coavariates ``gender of sender" and ``gender of receiver" are both significant in negative direction. \textbf{substantive interpretation on gender effects}. Next, the statistic ``outdegree" also has significantly negative effect---i.e., if sender $a$ sent $n$ number of emails to anyone last week, then sender $a$ is approximately $\exp(-0.109\times n)\approx(0.897)^n$ times less likely to send an email to $r$---possibly due to its multicollinearity with the statistics ``hyperedge size" where the two terms' show significant interaction effect. This result is likely to appear under the following scenario: if sender $a$ sent large number of emails over the last week but 1) not to $r$ (excluding the effect of ``send") and the email of interest is not a  2) broadcast email which is originally aimed at large number of receivers (excluding the effect of ``hyperedge size"), then sender $a$ is less likely to send an email to receiver $r$. Similar interpretation can be applied for the covariates ``indegree," but in the opposite direction. For example, if the receiver $r$ received $n$ number of emails over the last week, sender $a$ is $\exp(0.086\times n)\approx(1.091)^n $ times more likely to send an email to $r$.

	On the next plot for dyadic and triadic covariates, one thing we can check is that the effect of statistic``send" (i.e., number of times sender $a$ sent emails to receiver $r$ over the last week) is significant and positive with $\hat{b}_4 = 0.274$, implying that if sender $a$ sent $n$ number of emails to $r$ last week, then sender $a$ is approximately $\exp(0.274\times n)\approx(1.315)^n$ times more likely to send an email to $r$. Another dyadic statistic ``receive" is not statistically significant in that its 95\% credible interval do not include 0. Surprisingly, none of the covariates for triadic network effects seem to have significant effect on the generative process of edges.
					\caption{Nodal covariates}
					\includegraphics[width=\textwidth]{img/betanewplot2-1.png}	
					\caption{Dyadic and triadic covariates}
					\includegraphics[width=\textwidth]{img/betanewplot3-1.png}	
			\caption {Posterior distribution of $\boldsymbol{b}$ estimates.}
			\label{figure:betaresults}

@@

	To begin with, the posterior estimates of two temporal effects---``weekend" and 		AM/PM"---indicate that if the ${(d-1)}^{th}$ email was sent during the weekend or after midday, then the time to $d^{th}$ email is expected to take $\exp(1.552)\approx 4.722$ hours and $\exp(0.980)\approx2.665$ hours longer, respectively, compared to their counterparts (i.e., weekdays and am). On the contrary, the covariates ``manager", ``outdegree", and ``indegree" turn out to shorten the amount of time to next email. For example, being a county manager (i.e., the manager of department managers) lowers the expected value of $\log(\tau_{ad})$ by $\hat{\eta}_3 = -1.070$, since the manager in general sends or receives a lot more emails which may shorten the response time and many of those emails. This argument is also supported by the posterior estimates for ``outdegree" and ``indegree" statistics, where the estimated coefficients are approximately $\hat{\eta}_4=-0.206$ and $\hat{\eta}_5=-0.060$, respectively. Gender of the department manager is shown to have no significant effect on time increments. In addition, the posterior mean estimates for variance parameter $\sigma^2_\tau$ in lognormal distribution is approximately $\hat{\sigma}^2_\tau=14.093$ with its 95\% credible interval $(12.709, 15.555)$, indicating that there exists huge variability in the time increments of emails.

@@

@@

Motivated by the relational event model \citep{Butts2008} and its variants for time-stamped networks, the hyperedge event model (HEM) can effectively learn the underlying edge generating and timestamp generating structures of event data, while allowing flexibity in the covariate specifications and the distributional assumption of time increments.
Accounting for the hyperedges, we make better use of events with hyperedges than modeling them as duplicate edges. In modeling the timestamps of the events, our generalized linear model (GLM) based formulation eliminates the need to stick with one distribution (e.g., exponential distribution), and we also provide an algorithm for predictive experiment that helps to learn which distribution better fits the data. To sum, the HEM's flexible generative process thus leads to more accurate and precise inference on model parameters.

We have demonstrated effectiveness of our model by analyzing Montgomery County government emails, which includes numerous hyperedges. The estimated covariate effects reveal that the HEM is able to understand the structural dynamics similar to those used in the exponential random graph model (ERGM), while the model additionally allows separate covariates and the corresponding parameters for generative processes of edges and timestamps. Although we illustrate the entire framework and application in the context
of one type of hyperedges, a single sender and multiple receivers, our model can be easily extended to allow the opposite case, a single receiver and multiple senders, by slight modification of the generative process. This extension involves promissing applications including international sanctions and co-sponsorship of bills. Furthermore, while currently the model's applicability is limited to small-sized networks since we assume fixed set of nodes across entire timepoints, we can adjust the model to allow time-varying nodes or improve its computational efficiency such that the model is well applicable to large-scale networks. Considering the recent explosion of network dataset with huge number of nodes (e.g., online comnunications), the development of model adjustments for better applicability are objects of future work.

>>>>

	We evaluate the model's ability to predict edges and timestamps from Montgomery County email data, conditioned on their ``training" part of the data. To perform the experiment, we separately formed a test split of each three model components---sender, receivers, and timestamps---by randomly selecting ``test" data with probability $p=0.1$. Any missing variables were imputed by drawing samples from their conditional posterior distributions, given the observed data, model estimates, and current values of test data. We then run inference to update the latent variables given the imputed and observed data. We iterate the two steps---imputation and inference---multiple times to obtain enough number of estimates for test data. Algorithm \ref{alg:PPE} outlines this procedure in detail. We run the experiment and measure the predictive performance of two separate time distributions using $N=500$ predicted samples, by comparing the predictions in terms of classification accuracy in predicting the senders and receivers, as well as prediction error in the timestamps. We summarize the results of prediction experiments for missing senders, receivers, and timestamps in Figure \ref{figure:PPEresults}. First, we compare the posterior probability of correct senders for each of missing emails $\{d:a_d=\mbox{NA}\}$, which corresponds to $\pi_{a_{d}}=P(a_{d} = a^{obs}_{d}|\cdot)$ in Algorithm \ref{alg:PPE}. We call this measure as ``correct sender posterior probability." On Figure \ref{figure:PPEresults} (a), we draw boxplots for the distribution of mean correct sender posterior probability---i.e., $\hat{\pi}_{a_{d}} = \frac{1}{N}\sum_{n=1}^N \pi^n_{a_{d}}$---across the missing emails. The results show that both log-normal and exponential distributions acheive better predictive performance for missing senders compared to what is expected under random guess (i.e., choose one out of $A$ possible senders $=1/18$), with log-normal distribution showing slightly higher performance than exponential distribution on average. Secondly, since the receiver vector is binary, we compute $F_1$ scores for missing receiver indicators (i.e., all $d$ and $r$ with $r_{dr}$=NA) by taking the harmonic mean of precision and recall:

@@

	with TP denoting true positive (i.e., $\boldsymbol{r}^{obs}_{dr}=\boldsymbol{r}^{pred}_{dr}=1$), FN denoting false negative (i.e., $\boldsymbol{r}^{obs}_{dr}=1$ but $\boldsymbol{r}^{pred}_{dr}=0$), and FP denoting false positive (i.e., $\boldsymbol{r}^{obs}_{dr}=0$ but $\boldsymbol{r}^{pred}_{dr}=1$). Although the generative process for edges (Section \ref{subsec: Tie}) is not directly affected by the choice of timestamp distribution, Figure \ref{figure:PPEresults} (b) reveals noticeable difference between log-normal and exponential in their performance in predicting missing receiver indicators, where log-normal outperforms exponential. Finally, the prediction error for $d^{th}$ missing timestamp is estimated using the median of absolute relative errors\footnote{\cite{hyndman2006another} refer to this as median absolute percentage error (MdAPE).} across $N=500$ predictions:
			\begin{tabular}[t]{ccc}
				\begin{subfigure}[b]{0.33\textwidth}
					\caption{Sender prediction}
					\includegraphics[width=\textwidth]{img/senderpredict-1.png}	
				\begin{subfigure}[b]{0.33\textwidth}
					\caption{Receiver prediction}
					\includegraphics[width=\textwidth]{img/receiverpredict-1.png}	
				\begin{subfigure}[b]{0.33\textwidth}
					\caption{Timestamp prediction}
					\includegraphics[width=\textwidth]{img/timepredict-1.png}	
			\caption {Comparison of predictive performance between log-normal and exponential distributions: (a) correct sender posterior probability from sender predictions, (b) $F_1$ scores from receiver predictions, and (c) median absolute relative error from timestamp predictions. Blue line in (a) represents the correct sender probability expected by random guess---i.e., $1/A=1/18\approx0.056$.}
		\iffalse

@@

@@

		\fi

@@

	Figure \ref{figure:PPEresults} (c) presents boxplots for the median absolute relative errors. where plot the estimates in a log-scale. Surprisingly, we have a huge benefit in the performance of timestamp prediction when we assume log-normal distribution for time-increments compared to exponential distribution. This difference can be explained by overdispersion in exponential distribution, because there exists greater variability in the time increments of emails than would be expected under exponential distribution. As illustrated above, we can use this out-of-sample prediction task for two usage---1) provide an effective answer to the question ``how are we filling in the missing components of time-stamped network data?" and 2) offer one standard way to determine the distribution of time increments in Section \ref{subsec:Time}.

@@

	\subsubsection{Coefficients for edge covariates}
	Figure \ref{figure:betaresults} shows the boxplots summarizing posterior samples of $\boldsymbol{b}$, where Figure \ref{figure:betaresults} (a) displays the coefficients for nodal covariates and \ref{figure:betaresults} (b) displays the coefficients for dyadic and triadic covariates. Since we use the logit probability of $\lambda_{adr}$ when generating hypothetical edges in Section \ref{subsec: Tie}, we assume that 
	\begin{equation*}
	\mbox{logit}(\lambda_{adr})=\log\Big(\frac{\lambda_{adr}}{1-\lambda_{adr}}\Big) =b_{1}+b_{2} x_{adr2}\ldots+b_{14}x_{adr14},
	\end{equation*}
			\begin{figure}[!t]
				\centering
				\begin{tabular}[t]{cc}
					\begin{subfigure}[b]{0.495\textwidth}
						\caption{Nodal covariates}
						\includegraphics[width=\textwidth]{img/betanewplot2-1.png}	
					\end{subfigure}
					\begin{subfigure}[b]{0.495\textwidth}
						\caption{Dyadic and triadic covariates}
						\includegraphics[width=\textwidth]{img/betanewplot3-1.png}	
					\end{subfigure}
				\end{tabular}
				\caption {Posterior distribution of $\boldsymbol{b}$ estimates.}
				\label{figure:betaresults}
			\end{figure}
	and can interpret the $\boldsymbol{b}$ estimates in terms of odds ratio $\frac{\lambda_{adr}}{1-\lambda_{adr}}=\exp(b_{1}+b_{2} x_{adr2}\ldots+b_{14}x_{adr14})$.
	To begin with, we find out that the effect of nodal coavariates ``gender of sender" and ``gender of receiver" are both significant in negative direction. \textbf{substantive interpretation on gender effects}. Next, the statistic ``outdegree" also has significantly negative effect---i.e., if sender $a$ sent $n$ number of emails to anyone last week, then sender $a$ is approximately $\exp(-0.109\times n)\approx(0.897)^n$ times less likely to send an email to $r$---possibly due to its multicollinearity with the statistics ``hyperedge size" where the two terms' show significant interaction effect. This result is likely to appear under the following scenario: if sender $a$ sent large number of emails over the last week but 1) not to $r$ (excluding the effect of ``send") and the email of interest is not a  2) broadcast email which is originally aimed at large number of receivers (excluding the effect of ``hyperedge size"), then sender $a$ is less likely to send an email to receiver $r$. Similar interpretation can be applied for the covariates ``indegree," but in the opposite direction. For example, if the receiver $r$ received $n$ number of emails over the last week, sender $a$ is $\exp(0.086\times n)\approx(1.091)^n $ times more likely to send an email to $r$.

	When we look at the effects of dyadic and triadic covariates, one thing we can check is that the effect of statistic``send" (i.e., number of times sender $a$ sent emails to receiver $r$ over the last week) is significant and positive with $\hat{b}_4 = 0.274$, implying that if sender $a$ sent $n$ number of emails to $r$ last week, then sender $a$ is approximately $\exp(0.274\times n)\approx(1.315)^n$ times more likely to send an email to $r$. Another dyadic statistic ``receive" is not statistically significant in that its 95\% credible interval do not include 0. Surprisingly, none of the covariates for triadic network effects seem to have significant effect on the generative process of edges.

@@

@@

	To begin with, the posterior estimates of two temporal effects---``weekend" and ``PM"---indicate that if the ${(d-1)}^{th}$ email was sent during the weekend or after midday, then the time to $d^{th}$ email is expected to take $\exp(1.552)\approx 4.722$ hours and $\exp(0.980)\approx2.665$ hours longer, respectively, compared to their counterparts (i.e., weekdays and am). On the contrary, the covariates ``manager", ``outdegree", and ``indegree" turn out to shorten the amount of time to next email. For example, being a county manager (i.e., the manager of department managers) lowers the expected value of $\log(\tau_{ad})$ by $\hat{\eta}_3 = -1.070$, since the manager in general sends or receives a lot more emails which may shorten the response time and many of those emails. This argument is also supported by the posterior estimates for ``outdegree" and ``indegree" statistics, where the estimated coefficients are approximately $\hat{\eta}_4=-0.206$ and $\hat{\eta}_5=-0.060$, respectively. Gender of the department manager is shown to have no significant effect on time increments. In addition, the posterior mean estimates for variance parameter $\sigma^2_\tau$ in lognormal distribution is approximately $\hat{\sigma}^2_\tau=14.093$ with its 95\% credible interval $(12.709, 15.555)$, indicating that there exists large variability in the time increments of emails.

@@

Motivated by a growing class of dynamic network models which deal with edges exchanged in continuous time, the hyperedge event model (HEM) can effectively learn the underlying dynamics in edge and timestamp formations, providing novel insights to the literature. First, we explicitly model hyperedges by deriving a receiver-selection distribution that forces the sender to select at least one recevier, which is more realistic approach for events with one sender and one or more receivers and one or more sender and one receiver compared to treating them as pure duplicates. In modeling the timestamps (more precisely time increments) of events, our generalized linear model (GLM) based formulation offers new innovations by eliminating the need to stick with one parameter distribution (e.g., exponential distribution). To make better use of the proposed model, we provide an algorithm for predictive experiment that helps to learn which distribution better fits the data. To sum, the HEM's flexible generative process thus leads to more accurate and precise inference on model parameters.

We have demonstrated effectiveness of our model by analyzing Montgomery County government emails, where emails serve as the canonical example of directed edges with one sender and one or more receivers. The estimated effects for edge covariates reveal that the HEM is able to understand the structural dynamics similar to those used in the exponential random graph model (ERGM), but we also learn about the effect of timestamp covariates on waiting time to next event, by integrating a survival model for event timing. Although we illustrate the entire framework and application in the context of one type of hyperedges, one sender and one or more receivers, our model can be easily extended to allow the opposite case, one or more sender and one receiver, by slight modification of the generative process (shown in Appendix A). This extension involves promissing applications including international sanctions and co-sponsorship of bills. Furthermore, while currently the model's applicability is limited to small-sized networks since we assume fixed set of nodes across entire timepoints, we can adjust the model to allow time-varying nodes or improve its computational efficiency such that the model is well applicable to large-scale networks. Considering the recent explosion of network dataset with huge number of nodes (e.g., online comnunications), the development of model adjustments for better applicability are the object of future work.
